Cache is King: Presenting {\tt Corrfunc} -- a Suite of Fast Correlation Function Codes

### 2016-06-22 paper/ms.tex
latest paper changes

\documentclass[preprint,authoryear,12pt]{elsarticle}

@@

%%% Definitions added by Manodeep 
\newcommand{\Msun}{M_{\odot}}
\newcommand{\hMsun}{h^{-1}M_{\odot}}
\newcommand{\sdss}{{\small{SDSS}}\xspace}
\newcommand{\hMpc}{\ensuremath{{h^{-1}Mpc}\xspace}}
\newcommand{\hkpc}{\ensuremath{{h^{-1}kpc}\xspace}}
\newcommand{\emcee}{\texttt emcee}
\newcommand{\rmax}{\ensuremath{{r_{max}}}\xspace}
\newcommand{\xir}{\ensuremath{{DD(r)}}\xspace}
\newcommand{\xiofr}{\ensuremath{{\xi(r)}}\xspace}
\newcommand{\wprp}{\ensuremath{{w_p(r_p)}}\xspace}
\newcommand{\xirppi}{\ensuremath{{\xi(r_p,\pi)}}\xspace}
\newcommand{\todo}[1]{\marginpar{TODO}{\color{red}#1}}

@@

\title{Cache is King: Presenting a Suite of Fast, Correlation Function Codes}

@@

\begin{abstract}
We have entered the era of `Big Data' where we are gathering a tremendous amount of data that
needs to be processed as well as modeled accurately. In particular, large galaxy surveys
like the Sloan Digitial Sky Survey requires computing correlation functions. While measuring the
correlation function in the data is not a bottle-neck, modeling the observed galaxy distribution correctly
{\em} requires a MCMC chain and repeated measurements of $\xi(r)$ and/or $\xi(r_p,\pi)$. However, 
the architecture of modern CPU's means that best performance is only obtained when cache misses 
are kept to a minimum. Here, I present a suite of 3 correlation function codes that exploit 
current CPU architecture and hand-written AVX

The accompanying codes are publicly available at \url{https://bitbucket.org/manodeep/Corrfunc/}. 
\end{abstract}

@@

\section{INTRODUCTION}
%% %% MS: I have not encountered this in the past -- but the footnotecounter needs to be reset. Otherwise, 
%% %% the numbering starts at Nauthors+1. Presumably \altaffiltext uses \footnote internally and does not
%% %% reset the counter like a responsible person!
%% \setcounter{footnote}{0}
The large-scale structure of the Universe can be now measured using $\mathcal{O}$(million) galaxies 
using data from current surveys like SDSS/BOSS surveys. Upcoming surveys like LSST will probe even 
deeper and wider and aim to target on $\mathcal{O}$(10's of million) galaxies. With such a large galaxy data, 
we can measure the galaxy density field fairly accurately in the data. While the number of galaxies observed 
in these current and upcoming surveys is large, we have to compute the galaxy density fields {\em only once}. Thus, 
even a slow, correlation function code will be fine when measuring the data correlation function. 
However, predicting this galaxy  density field data will require an even larger number of model galaxies -- increasing the computational 
load of determining the density field even further. Typically, the modeling process also involves an MCMC -- 
requiring $\sim$ millions of evaluations of the correlation function.\footnote{I will assume that creating the models themselves 
is much faster compared to computing the correlation functions.}

Modern cpu's have a hierarchy of memory locations; the smallest (and fastest) are closest to the cores while the largest (and slowest) are the
farthest. All cpu instructions need to be carried out from cpu registers - there are $\sim 10-30$ registers typically available and the 
access times can be thought of as instantaneous. Next up is the $L1$ cache divided into $L1D$ for data and $L1I$ for instructions cache. Since 
the cpu always necessarily executes instructions that are close together, we will ignore the instruction cache from now on. Typical $L1$ cache 
sizes range from 64KB to 128 KB. Next level up is the $L2$ cache, typically $\sim$ 256KB to 1 MB. The last level cache or the $L3$ cache 
is usually shared across all cores on the socket and can be 10 MB to 40 MB.

>>>>

\documentclass[preprint, 12pt, authoryear]{elsarticle}

@@

%%% Definitions added by Manodeep 
\newcommand{\Msun}{M_{\odot}}
\newcommand{\hMsun}{h^{-1}M_{\odot}}
\newcommand{\sdss}{{\small{SDSS}}\xspace}
\newcommand{\hMpc}{\ensuremath{{h^{-1}Mpc}\xspace}}
\newcommand{\hkpc}{\ensuremath{{h^{-1}kpc}\xspace}}
\newcommand{\emcee}{\texttt emcee}
\newcommand{\rmax}{\ensuremath{{r_{max}}}\xspace}
\newcommand{\xir}{\ensuremath{{DD(r)}}\xspace}
\newcommand{\xiofr}{\ensuremath{{\xi(r)}}\xspace}
\newcommand{\wprp}{\ensuremath{{w_p(r_p)}}\xspace}
\newcommand{\xirppi}{\ensuremath{{\xi(r_p,\pi)}}\xspace}
\newcommand{\cis}{\ensuremath{{pN(r)}}\xspace}
\newcommand{\ddrppi}{\ensuremath{{DD(r_p,\pi)}}\xspace}
\newcommand{\wtheta}{\ensuremath{{DD(\theta)}}\xspace}
\newcommand{\todo}[1]{\marginpar{TODO}{\color{red}#1}}

@@

\title{Cache is King: Presenting {\tt Corrfunc} -- a Suite of Fast Correlation Function Codes}

@@

\begin{abstract}
In the era of `Big Data', we need special computing techniques to both process the raw data as well 
generate and evaluate models to test our theories. In particular, large galaxy surveys
like the Sloan Digitial Sky Survey requires computing a variety of $n$-point statistics, e.g., the 
2-pt projected correlation function, $\wprp$. 
Since the clustering measure on the data is only done once, time taken to do
such a measurement is not a bottleneck. However, modeling the observed galaxy distribution correctly
{\em requires} a Monte-Carlo Markov Chain (MCMC) and repeated measurements of one or multiple
clustering statistics. In order to efficiently compute {\em any} clustering statistic,
we need to write code that is tuned for the underlying hardware. For
modern CPUs, such tuning involves proper utilization of the cache hierarchy {\em and}
vectorization with Single Instruction Multiple Data (SIMD) capable wide vector registers. 
In this paper, I present {\tt Corrfunc} --  a suite of OpenMP parallelized, clustering
codes that exploit current CPU micro-architecture with hand-written Advanced
Vector Extensions (AVX) and  Streaming SIMD Extensions (SSE) intrinsics. 
Given a 3-D distribution of points in a cartesian system (i.e.,
generated from cosmological simulations), {\tt Corrfunc} can compute spatial
3-D correlation functions \xir, \xiofr, \xirppi; 2-point 
projected correlation function \wprp; counts-in-spheres \cis. For a 3-D
particle distribution on the sky (i.e., mock galaxy catalogs, observed galaxies), the codes can compute 
projected correlation function \ddrppi, angular correlation function \wtheta as
well as \cis.  By design, {\tt Corrfunc} is highly optimized 
and can compute \wprp for $\mathcal{O}$(1 million) galaxies in
$\sim 6$ seconds on a post-2011 CPU, which is a factor of few faster than 
existing public correlation function routines.  {\tt Corrfunc} is
publicly available at \url{https://github.com/manodeep/Corrfunc/} while
benchmarks with existing codes are available at
\url{https://gist.github.com/manodeep/cffd9a5d77510e43ccf0}.
\end{abstract}

@@

\section{INTRODUCTION}
The large-scale structure of the Universe can be now measured using $\mathcal{O}$(million) galaxies 
using data from current surveys like the Sloan Digitial Sky Survey (SDSS). Upcoming
surveys on the Large Synoptic Survey Telescope will probe even 
deeper and wider and aim to target on $\mathcal{O}$(10's of million) galaxies. With such a large galaxy data, 
we can measure the galaxy density field fairly accurately in the data. While the number of galaxies observed 
in these current and upcoming surveys is large, we have to compute the galaxy density fields {\em only once}. Thus, 
even a slow, correlation function code will be fine when measuring the data correlation function. 
However, predicting this galaxy  density field data will require an even larger number of model galaxies -- increasing the computational 
load of determining the density field even further. Typically, the modeling process also involves an MCMC -- 
requiring $\sim$ millions of evaluations of the correlation function.\footnote{I will assume that creating the models themselves 
is much faster compared to computing the correlation functions.}

\section{BACKGROUND}
\subsection{HARDWARE ARCHITECTURE}
Modern cpu's have a hierarchy of memory locations; the smallest (and fastest) are physically located close to the computing cores 
while the largest (and slowest) are the farthest. All cpu instructions need to be carried out from cpu registers - there are $\sim 100$ 
registers typically available and the access times are essentially
instantaneous. Next up is the {\em L1} cache 
divided into {\em L1D} for data and {\em L1I} for instructions cache. Since 
the cpu always necessarily executes instructions that are close together, we will ignore the instruction cache from now on. Typical L1 cache 
sizes range from 64KB to 128 KB (shared between instruction and data). 
Next level up is the L2 cache, typically $\sim$ 256KB to 1 MB. The last level cache or the L3 cache is usually shared across all 
cores on the socket and can be 10 MB to 40 MB. 
\subsection{VECTORIZATION}

\subsection{OPTIMIZATION STRATEGIES}

\section{PACKAGE SUMMARY}
{\tt Corrfunc} is written primarily in {\tt C} and comes with convenient
{\tt python} wrappers for every clustering statistic. The fundamental design
principles for {\tt Corrfunc} are the following:
\begin{enumerate}
  \item Correct -- {\tt Corrfunc} has a base set of correct outputs for every statistic
    generated either through slow, brute-force methods or independent, 
    external codes. Within {\tt Corrfunc}, every clustering statistic is covered with at least one
    test case that {\em requires} reproducing this `known-correct' result {\em
      exactly}. The `PASS/FAIL' status for each test is the direct output of {\tt diff -q
      test\_output known\_correct\_output}.  
  \item High Performance -- Performance is the overarching goal for {\tt
    Corrfunc}. The code-base has been repeatedly analyzed with Intel VTune
    (\url{https://software.intel.com/en-us/intel-vtune-amplifier-xe})
    on a variety of platforms to optimize the innermost loops. 
  \item Portable -- {\tt Corrfunc} is written in {\tt ISO/IEC 9899:1999}
    compliant {\tt C}. All instruction set specific codes are protected via compile-time constant
    definitions.
\end{enumerate}
Every clustering statistic in {\tt Corrfunc} can either be accessed through the
API call (e.g., via {\tt python}), linking with the static libraries) or as
a command-line executable.
